Topological Equivariant Artist Model

### 2021-03-17 draft/sections/math.tex
cleaned up fibers

We model the data $\mathscr{\dtotal}$, graphic $\mathscr{\gtotal}$, and intermediate visual encoding $\mathscr{\vtotal}$ stages of visualization as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic. To explain which structure the artist is preserving, we first describe how we model data (\ref{sec:data}), graphics (\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}) as fiber bundles. We then discuss the equivariant maps between data and visual characteristics (\ref{sec:artist_nu}) and visual characteristics and graphics (\ref{sec:artist_q}) that make up the artist.

>>>>

We model the data $\mathscr{\dtotal}$, graphic $\mathscr{\gtotal}$, and intermediate visual encoding $\mathscr{\vtotal}$ stages of visualization as topological structures that encapsulate types of variables and continuity. To explain which structure the artist is preserving, we first describe how we model data (\ref{sec:data}), graphics (\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}) as fiber bundles. We then discuss the equivariant maps between data and visual characteristics (\ref{sec:artist_nu}) and visual characteristics and graphics (\ref{sec:artist_q}) that make up the artist.

### 2021-03-17 paper/paper.tex
cleaned up fibers

\section{Topological Equivariant Artist Model}
\label{sec:math}
We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map
\begin{equation}
    \label{eq:artist}
    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
\end{equation}
from data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$. To formalize the structure the artist is preserving, we describe how we model data (\ref{sec:math:data}) and graphics (\ref{sec:math:graphic}) as fiber bundles. We then discuss the equivariant maps that make up $\mathscr{\vartist}$, which are a map from  graphic continuity to data continuity (\ref{sec:math:graphic:base}), data components to visual components (\ref{sec:math:artist:nu}) and visual components to graphic (\ref{sec:math:artist:q}).

\subsection{Data Bundle}
\label{sec:math:data}
Building on Butler's proposal of using fiber bundles as a common data representation structure for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}, a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
\begin{equation}
    \label{eq:fiber_bundle}
    \begin{tikzcd}
        \dfiber \arrow[r, hook] & \dtotal \arrow[r, "\pi"] & \dbase
    \end{tikzcd}
\end{equation}
that binds the components of the data in \dfiber\ to the continuity represented in \dbase. By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood $U$ the total space is the cartestian product $\dbase\times\dfiber$.

\subsubsection{Fiber Space: Variables}
\label{sec:math:data:fiber}
\begin{figure}[htb]
  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
  \includegraphics[width=\columnwidth]{fiber.png}
  \caption{These two datasets have the same base space \dbase\ of discrete points, but the square is a visual representation of the fiber space  $\dfiber=\reals\times\reals$, while the cube is a representation of $\dfiber=\reals\times\realsp\times\reals$}
  \label{fig:math:data:fiber}
\end{figure}
To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names. Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset. He then defines the single variable set \fttype\ 
\begin{equation}
    \label{eq:data_types}
    \begin{tikzcd}
        \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] & \ftotal \arrow[d, "\pi"] \\
        \fnames \arrow[r, "\fsection"']                          & \ftypes       
    \end{tikzcd}
\end{equation}
which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname. The \fttype\ lookup is by name to specify that every component is distinct, since multiple components can have the same type \ftype. Given \fsection, the fiber for a one variable dataset is
\begin{equation}
    \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
\end{equation}
where \fsection\ is the schema binding variable name \fname\ to its datatype \ftype. A dataset with multiple variables has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
\begin{equation}
F = \ftotal_{\fsection(\fname_{1})}\times \ldots \ftotal_{\fsection(\fname_{i})} \ldots\times \ftotal_{\fsection(\fname_{n})}
\end{equation}
which is equivalent to 
\begin{equation}
    \label{eq:math:data:fiber:decompose}
    \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
\end{equation}
which allows us to decouple \dfiber\ into components $\dfiber_i$. Each component of \dfiber\ is a dimension of the topological fiber space, as illustrated in \autoref{fig:math:data:fiber}. The space $\dfiber=\reals\times\reals$ encodes that the dataset has two quantative continuous variables, while $\dfiber=\reals\times\reals\realsp$ encodes a dataset with three quantative contunous variables, one of which must be positive.

@@

A monoid \cite{Monoid2021} $\monoid$ is a set with an associative binary operator $\ast:\monoid \times \monoid\rightarrow \monoid$. A monoid has an identity element $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
\begin{align*}
    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
\end{align*}
As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
\begin{equation}
\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
\end{equation}
of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$

@@

\subsubsection{Base Space: Continuity}
\label{sec:math:data:base}
 The base space \dbase\ is way to express how the records in \dtotal\ are connected to each other, for example if they are discrete points or if they lie in a 2D continous surface. Formally \dbase\ is the quotient space of \dtotal\, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$\cite{QuotientSpaceTopology2020}. Our model formulates \dbase\ as akin to an indexing space into \dtotal\ that describes the structure of \dtotal.

@@

 As illustrated in ~\autoref{fig:math:data:base}, \dbase\ can have any number of dimensions, can be continuous or discrete, and is somewhat independent of the dimensions of the fiber. As with fibers and monoids, we can decompose the total space into components $\pi:\dtotal_i\rightarrow \dbase$ where
 \begin{equation}
    \label{eq:math:data:base:decompose}
     \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
 \end{equation}
 which is a decomposition of \dfiber. The \dbase\ remains the same because the connectivity of records does not change just because there are fewer elements in each record. By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit. The explicit topology is a concise way of distinguishing visualizations that appear identical, for example heatmaps and images.

@@

 \subsection{Graphic Bundle}
 \label{sec:math:graphic}  
We introduce a graphic bundle to hold the essential information necessary to render a graphical design constructed by the artist. As with the data, we can represent the target graphic as a section \gsection\ of a bundle  $(\gtotal, \gbase, \pi, \gfiber)$
\begin{equation}
  \begin{tikzcd}[ampersand replacement=\&]
      \gfiber \arrow[r, hook] \& \gtotal \arrow[d, "\pi"'] \\
                        \& \gbase \arrow[u, "\gsection"', bend right]
  \end{tikzcd}
\end{equation}
where \gsection\ is the fully encoded graphic. To fully specify the visual characteristics of the image, we construct a fiber \gfiber\ that is an infinite resolution version of the target space. Typically \gtotal\ is trivial and therefore sections can be thought of as mappings into \gfiber. In this work, we assume a 2D opaque image $\gfiber=\reals^5$ with elements $(x,\, y,\, r,\, g,\, b) \in \gfiber$ such that a rendered graphic only consists of 2D position and color. By abstracting the target display space as \gfiber, the model can support different targets, such as a 2D screen or 3D printer.

@@

Just as the \dbase\ encodes the connectivity of the records in the data, we propose an equivalent \gbase\ that encodes the connectivity of the rendered elements of the graphic. Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ so that \dbase\ and \gbase\ have the same homotopy. The surjective map $\vindex: \gbase \rightarrow \dbase$ 
\begin{equation}
    \begin{tikzcd}
        \dtotal \arrow[d, "\pi"'] & \gtotal \arrow[d, "\pi"'] \\
        \dbase                   & \gbase \arrow[l, "\vindex"']
    \end{tikzcd}
\end{equation}
goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$. While \gbase\ must have the same continuity as \dbase\, it is sometimes the  thickened version shown in \autoref{fig:math:graphic:base}. This thickening is necessary when the dimensionality of \dbase\ is less than the dimensionality of the target display. For example, a \dbasepoint\ that is 0D in \dbase\ cannot be represented on screen unless it is thickened to 2D to encode the connectivity of the points in \gfiber\ that visually represent the record at \dbasepoint.

\subsection{Artist}
The topological artist \vartist\ is a monoid equivariant sheaf map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$. 
\begin{equation}
    \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
\end{equation}
While \vartist\ can usually construct graphical elements solely with the data in \dsection, some visualizations, such as line, may also need some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$. In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives; therefore the artist takes as input the jet bundle $\dtotal^{\prime}=\mathcal{J}^{2}(\dtotal)$.

Specifically, \vartist\ is the equivariant map from $\dtotal^{\prime}$ to a specific graphic $\gsection \in \Gamma(H)$ 
\begin{equation}
    \label{eq:math:artist:diagram}
    \begin{tikzcd}
        \dtotal^{\prime} \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] & \vtotal \arrow[d, "\pi"] & \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] & \gtotal \arrow[ld, "\pi"] \\
                                              & \dbase                  & \gbase \arrow[l, "\vindex"']                                              &                    
        \end{tikzcd}
\end{equation}
where the input can be point wise $\dsection(\dbasepoint)\mid \dbasepoint \in \dbase$. The visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ is the latent space of possible parameters of a visualization type, such as a scatter or line plot. As with the data and graphic bundles, the visual bundle is  defined by the projection map $\pi$
\begin{equation}
    \begin{tikzcd}[ampersand replacement=\&]
        \vfiber \arrow[r, hook] \& \vtotal \arrow[d, "\pi"'] \\
                          \& \dbase \arrow[u, "\vsection"', bend right]
    \end{tikzcd}
\end{equation}
where \vsection\ is the visual variable encoding\cite{bertinSemiologyGraphicsDiagrams2011a} of the data section \dsection. The visual fiber \vfiber\ is defined in terms of the input parameters of the visualization library's plotting functions; by making these parameters explicit components of the fiber, we can build consistent definitions and expectations of how these parameters behave.

In \autoref{eq:math:artist:diagram}, the encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components. The continuity map $\vindex:\gbase \rightarrow \dbase$ then pulls back the visual bundle \vtotal\ over \gbase. Then the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal. This functional decomposition of the visualization artist facilitates building reusable components at each stage of the transformation because the equivariance constraints are defined on \vchannel, \vmark, and \vindex. We name this map the artist as that is the analogous part of the  Matplotlib\cite{hunterArchitectureOpenSource} architecture that builds visual elements.

@@

In \autoref{fig:teaser}, we illustrate the output of a minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image. The scatter plot can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase. The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in \autoref{fig:math:artist:base}
\begin{align*}
x &= size *\alpha \cos(\beta) + xpos \\
y &= size *\alpha \sin(\beta) + ypos
\end{align*}
such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black. In contrast, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ in \autoref{fig:teaser} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase. As shown in \ref{fig:math:arist:base} line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness. The magnitude of the slope is $\lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}$
such that the normal is  $\hat{n_{1}} = \frac{n_1}{\lvert n \rvert}, \; \hat{n_{2}} = \frac{n_2}{\lvert n \rvert}$ which yields components of \gsection\
\begin{align*}
 x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n_1}(\vindex(\alpha)) \\
 y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n_2}(\vindex(\alpha)) 
\end{align*}
where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n_1}, \hat{n_2}$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.

@@

\subsubsection{Assembly Factory}
\label{sec:math:artis:qhat}
The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark. As shown in eq~\ref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
\begin{figure}[htb]
  \centering
    \includegraphics[width=1\columnwidth]{q_hat.png}
    \caption{Because the pullback of the visual bundle \vtotalpull\ is the replication of a \vsection\ over all points \gbasepoint\ that map back to a single \dbasepoint, we can constract a \vmarkd\ on \vsection\ over \dbasepoint\ that will fabricate the \vmark\ for the equivalent region of \gbasepoint\ associated to that \dbasepoint}
    \label{fig:math:artist:qhat}
\end{figure}

@@

Factoring out \gbasepoint\ from equation~\ref{eq:math:artist:qhat_q_s} yields $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase. In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record  $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$. This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$. The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing pipeline for libraries that delegate the construction of \gsection\ to a back end such as Matplotlib.

>>>>

\section{Topological Equivariant Artist Model}
\label{sec:math}
We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map
\begin{equation}
    \label{eq:artist}
    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
\end{equation}
from data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$. To formalize the structure the artist is preserving, we describe how we model data (\autoref{sec:math:data}) and graphics (\autoref{sec:math:graphic}) as fiber bundles. We then discuss the equivariant maps that make up $\mathscr{\vartist}$, which are a map from  graphic continuity to data continuity (\autoref{sec:math:graphic:base}), data components to visual components (\autoref{sec:math:artist:nu}) and visual components to graphic (\autoref{sec:math:artist:q}).

\subsection{Data Bundle}
\label{sec:math:data}
Building on Butler's proposal of using fiber bundles as a common data representation structure for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}, a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
\begin{equation}
    \label{eq:fiber_bundle}
    \begin{tikzcd}
        \dfiber \arrow[r, hook] & \dtotal \arrow[r, "\pi"] & \dbase
    \end{tikzcd}
\end{equation}
that binds the components of the data in \dfiber\ to the continuity represented in \dbase. By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood $U$ the total space is the cartesian product $\dbase\times\dfiber$.

\subsubsection{Fiber Space: Variables}
\label{sec:math:data:fiber}
To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names. Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset. He then defines the single variable set \fttype\ 
\begin{equation}
    \label{eq:data_types}
    \begin{tikzcd}
        \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] & \ftotal \arrow[d, "\pi"] \\
        \fnames \arrow[r, "\fsection"']                          & \ftypes       
    \end{tikzcd}
\end{equation}
which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname. The \fttype\ lookup is by name to specify that every component is distinct, since multiple components can have the same type \ftype. Given \fsection, the fiber for a one variable dataset is
\begin{equation}
    \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
\end{equation}
where \fsection\ is the schema binding variable name \fname\ to its datatype \ftype. A dataset with multiple variables has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
\begin{equation}
F = \ftotal_{\fsection(\fname_{1})}\times \ldots \ftotal_{\fsection(\fname_{i})} \ldots\times \ftotal_{\fsection(\fname_{n})}
\end{equation}
which is equivalent to 
\begin{equation}
    \label{eq:math:data:fiber:decompose}
    \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
\end{equation}
which allows us to decouple \dfiber\ into components $\dfiber_i$. Each component of \dfiber\ is a dimension of the topological fiber space and is specified by a tuple of the form (\fname, \ftype, \(\ftotal_{\fsection(\fname)}\)). 
\begin{figure}[htb]
  \centering 
  \includegraphics[width=\columnwidth]{fiber.png}
  \caption{These two datasets have the same base space \dbase. The plane is a representation of the fiber $\dfiber=\reals\times\reals$ for the variables (time, temperature),  while the cube is the fiber $\reals\times\realsp\times\reals$ associated with (time, wind=(speed, direction))}
  \label{fig:math:data:fiber}
\end{figure}
In \autoref{fig:math:data:fiber} the plane fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{temperature}, \texttt{float}, \reals), while the cube fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{wind}, \texttt{wind}, \(\realsp\times\reals)\)) which encodes (\emph{speed}, \emph{direction}).

@@

A monoid \cite{Monoid2021} $\monoid$ is a set with an associative binary operator $\ast:\monoid \times \monoid\rightarrow \monoid$. A monoid has an identity element $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
\begin{align*}
    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
\end{align*}
As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
\begin{equation}\label{eq:math:data:monoid:decompose}
\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
\end{equation}
of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$

@@

\subsubsection{Base Space: Continuity}
\label{sec:math:data:base}
 The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other and does not itself store any of the components of the data. For example, \dbase\ can encode that the timeseries data described in \autoref{sec:math:data:fiber} is continuous, but is doing so independent of the timestamps. Every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$ because \dbase\ is the quotient space\cite{QuotientSpaceTopology2020,aurouxMath131Introduction} of \dtotal.

@@

 As illustrated in ~\autoref{fig:math:data:base}, \dbase\ can have any number of dimensions, can be continuous or discrete, and is somewhat independent of the dimensions of the fiber. As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:monoid:decompose},  we can decompose the total space into component bundles
 \begin{equation}\label{eq:math:data:base:decompose}
     \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
 \end{equation}
 such that \(\monoid_{i}\) acts on component bundle \(\dtotal_i\). The \dbase\ remains the same because the connectivity of records does not change just because there are fewer components in each record. By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit. The explicit topology is a concise way of distinguishing visualizations that appear identical, for example heatmaps and images.

@@

 \subsection{Graphic Bundle}
 \label{sec:math:graphic}  
We introduce a graphic bundle to hold the essential information necessary to render a graphical design constructed by the artist. As with the data, we can represent the target graphic as a section \gsection\ of a bundle  $(\gtotal, \gbase, \pi, \gfiber)$
\begin{equation}
  \begin{tikzcd}[ampersand replacement=\&]
      \gfiber \arrow[r, hook] \& \gtotal \arrow[d, "\pi"'] \\
                        \& \gbase \arrow[u, "\gsection"', bend right]
  \end{tikzcd}
\end{equation}
where \gsection\ is a fully specified graphic such that it is an abstraction of rendering. To fully specify the visual characteristics of the image, we construct a fiber \gfiber\ that is an infinite resolution version of the target space. Typically \gtotal\ is trivial and therefore sections can be thought of as mappings into \gfiber. In this work, we assume a 2D opaque image $\gfiber=\reals^5$ with elements $(x,\, y,\, r,\, g,\, b) \in \gfiber$ such that a rendered graphic only consists of 2D position and color. By abstracting the target display space as \gfiber, the model can support different targets, such as a 2D screen or 3D printer.

@@

Just as the \dbase\ encodes the connectivity of the records in the data, we propose an equivalent \gbase\ that encodes the connectivity of the rendered elements of the graphic. Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ so that \dbase\ and \gbase\ have the same homotopy. The surjective map $\vindex: \gbase \rightarrow \dbase$ 
\begin{equation}
    \begin{tikzcd}
        \dtotal \arrow[d, "\pi"'] & \gtotal \arrow[d, "\pi"'] \\
        \dbase                   & \gbase \arrow[l, "\vindex"']
    \end{tikzcd}
\end{equation}
goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$. While \gbase\ must have the same continuity as \dbase\, it is sometimes the  thickened version shown in \autoref{fig:math:graphic:base}. This thickening is necessary when the dimensionality of \dbase\ is less than the dimensionality of the target display. For example, a \dbasepoint\ that is 0D in \dbase\ cannot be represented on screen unless it is thickened to 2D to encode the connectivity of the points in \gfiber\ that visually represent the record at \dbasepoint. The \vindex\ mapping is critical to interactive visualizations as it is the map from a region on screen to the data associated with that region. One example is to fill in details in a hover tooltip, another is to convert region selection on \gbase\ to a query on the data to access the corresponding record components on \dbase.

\subsection{Artist}
The topological artist \vartist\ is a map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$. 
\begin{equation}
    \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
\end{equation}
that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}. Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps 
\begin{equation}
\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
\end{equation}
such that applying a monoid action $m \in \monoid$ to the data $\delement \in \mathscr{\dtotal}$ input to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.

The monoid equivariant map has two stages: the encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components, and the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal.
\begin{equation}
    \label{eq:math:artist:diagram}
    \begin{tikzcd}
        \dtotal \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] & \vtotal \arrow[d, "\pi"] & \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] & \gtotal \arrow[ld, "\pi"] \\
                                              & \dbase                  & \gbase \arrow[l, "\vindex"']                                              &                    
        \end{tikzcd}
\end{equation}
\vtotalpull\ is the visual bundle \vtotal\ pulled back over \gbase\ via the equivariant continuity map $\vindex:\gbase \rightarrow \dbase$ introduced in \autoref{sec:math:graphic:base}. The visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ is the latent space of possible parameters of a visualization type, such as a scatter or line plot. As with the data and graphic bundles, the visual bundle is defined by the projection map $\pi$
\begin{equation}
    \begin{tikzcd}[ampersand replacement=\&]
        \vfiber \arrow[r, hook] \& \vtotal \arrow[d, "\pi"'] \\
                          \& \dbase \arrow[u, "\vsection"', bend right]
    \end{tikzcd}
\end{equation}
where \vsection\ is the visual variable encoding\cite{bertinSemiologyGraphicsDiagrams2011a} of the data section \dsection. The visual fiber \vfiber\ is defined in terms of the input parameters of the visualization library's plotting functions; by making these parameters explicit components of the fiber, we can build consistent definitions and expectations of how these parameters behave. The functional decomposition of the visualization artist facilitates building reusable components at each stage of the transformation because the equivariance constraints are defined on \vchannel, \vmark, and \vindex. We name this map the artist as that is the analogous part of the  Matplotlib\cite{hunterArchitectureOpenSource} architecture that builds visual elements.

@@

In \autoref{fig:teaser}, we illustrate the output of a minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image. The scatter plot can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase. The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in \autoref{fig:math:artist:base}
\begin{align*}
x &= size *\alpha \cos(\beta) + xpos \\
y &= size *\alpha \sin(\beta) + ypos
\end{align*}
such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black. In contrast, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ in \autoref{fig:teaser} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase. As shown in \autoref{fig:math:artist:base}, line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness; therefore the artist takes as input the jet bundle \cite{JetBundle2020,musilovaCalculusVariationsJet2016} $\mathcal{J}^{2}(\dtotal)$ which is the data \dtotal\ and the first and second derivatives of \dtotal. The magnitude of the slope is $\lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}$
such that the normal is  $\hat{n_{1}} = \frac{n_1}{\lvert n \rvert}, \; \hat{n_{2}} = \frac{n_2}{\lvert n \rvert}$ which yields components of \gsection\
\begin{align*}
 x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n_1}(\vindex(\alpha)) \\
 y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n_2}(\vindex(\alpha)) 
\end{align*}
where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n_1}, \hat{n_2}$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.

@@

\subsubsection{Assembly Factory}
\label{sec:math:artist:qhat}
The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark. As shown in \autoref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
\begin{figure}[htb]
  \centering
    \includegraphics[width=1\columnwidth]{q_hat.png}
    \caption{Because the pullback of the visual bundle \vtotalpull\ is the replication of a \vsection\ over all points \gbasepoint\ that map back to a single \dbasepoint, we can constract a \vmarkd\ on \vsection\ over \dbasepoint\ that will fabricate the \vmark\ for the equivalent region of \gbasepoint\ associated to that \dbasepoint}
    \label{fig:math:artist:qhat}
\end{figure}

@@

Factoring out \gbasepoint\ from \autoref{eq:math:artist:qhat_q_s} yields $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase. In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record  $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$. This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$. The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing pipeline for libraries that delegate the construction of \gsection\ to a back end such as Matplotlib.
